Embedding Intuitionistic into Classical Logic

### 2022-10-13 FoSSaCS/fossacs-EIICL.tex
typos

A detailed account of the translation can be found in definition~\ref{def:fo-translation}. Finally, we benchmark our translation by translating all problems in the ILTP problem set~\cite{iltp} and applying the Vampire theorem prover~\cite{kovacs2013first} to the translated instances. While this approach falls short of the absolute state of the art in general, under the right circumstances it is a reasonable approach. Since there is still a lot of room for improvement this is a hopeful first sign.

@@

\noindent As all we have done is modeling Kripke frames as a first-order theory so we directly obtain:

@@

\begin{remark}
	It is well known that Kripke frames which are trees are complete with regard to intuitionistic logic. Our contribution is giving an easy to compute explicit size and geometry that depends only on the structure of the considered formula.
\end{remark}

@@

We can apply the previous results to give a direct embedding from intuitionistic into classical propositional logic. We know that $\varphi^{CH}$ is valid iff it is valid for structures that have as a domain the sequences without repetition over $\mathcal F_\to$, ordered by the prefix-relation and $$s^I = \epsilon, f_\psi^I(x) = \begin{cases}
	x, &\text{ if $\psi$ occurs in $x$,}\\
	x\psi, &\text{ otherwise.}
\end{cases}$$
Now we can replace every $\forall$-quantifier in $\varphi^{CH}$ by enumerating all the ground terms, and then replace all distinct ground instances of relations by new propositional variables yielding a propositional formula. Note that the encoding of partial order becomes redundant with this so we drop it, but we need to keep the persistency constraints. This gives us the following theorem:

\begin{theorem}\label{thm:prop-translation}
	Suppose we are given a formula as in Lemma~\ref{lemma:propositional-normal-form}, i.e. of the form
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal S\right)\to P$$
	with flat $\mathcal R$ and implication $\mathcal S$. Let $\Lambda$ denote the set of sequences without repetition over $\mathcal X$. For every propositional variable $A$ and $\lambda\in\Lambda$ consider a new propositional Variable $A^\lambda$. Define $R^\#$ as the union over all sets
	$$
		\left\{\bigwedge_iA_i^\lambda\to\bigvee_jB_j^\lambda\:|\:\lambda\in\Lambda\right\}
	$$
	for $\bigwedge_iA_i\to\bigvee_jB_j\in\mathcal R$ and $\mathcal X^\#$ as the union over all sets
	$$
		\left\{\left(A^{\lambda\psi} \to A^{\lambda\psi}\right)\to(C^\lambda)\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	for $\psi = (A\to B)\to C\in\mathcal X$ as well as $P^\#$ as the union over all sets
	$$
		\left\{A^\lambda\to A^{\lambda\psi}\:|\:\lambda\in\Lambda, \psi\psi\text{ does not occur in }\lambda\right\}
	$$
	 for each propositional variable $A$ occurring in $\varphi$ encoding persistency. Then
	$$
		\varphi^\# := \left(\bigwedge P^\#\wedge\bigwedge \mathcal R^\#\wedge\bigwedge\mathcal X^\#\right)\to P^\epsilon
	$$
	is classically valid if and only if $\varphi$ is intuitionistically valid.
\end{theorem}
%\begin{theorem}
%	$\bigwedge\mathcal S\to P$ is intuitionistically valid iff $\bigwedge\mathcal S^\#\to P^\emptyset$ is classically valid.
%\end{theorem}

@@

\begin{definition}
	For every propositional variable $A$ occurring in $\varphi$ and non-negative integer $n$ consider a new propositional variable $A^n$ and for every formula $\psi\in\mathcal X$ consider a new propositional variable $X_\psi^n$. Let $\vec A^n$ range over all propositional variables $A^n$ and $\vec X^n$ range over all $X_\psi^n$. Define
	\begin{itemize}
		\item {\normalfont Valid}$(n)$ which encodes that $\vec X^n_\psi$ represents a valid next element of a sequence without repetition, i.e. if we interpret the $\vec X^i$ as bit-vectors, then $\vec X^n$ has exactly one bit set to $1$, indicating the formula at the $n$-th position of the sequence, and it is at a position that is $0$ in $\sum\{\vec X^i\:|\:i < n\}$.
		\item {\normalfont Persistent}$(n)$ which encodes that the persistency condition holds.
		\item {\normalfont Sat}$_{\mathcal R}(n)$ encoding that the formulas in $\mathcal R$ hold at the $n$-th world of the path.
		\item {\normalfont Sat}$_{\mathcal X}(n)$ encoding that the formula in $\mathcal X$ that is represented by $\vec X^{n-1}$ holds.
	\end{itemize}
	Have $k = |\mathcal X|$ and define$$\varphi^Q_{i} = \exists \vec A^i\forall \vec X^i\left(\text{\normalfont Persistent}(i)\wedge \text{\normalfont Sat}_{\mathcal R}(i)\wedge \text{\normalfont Sat}_{\mathcal X}(i)\wedge\left(\text{\normalfont Valid}(i)\to \varphi^Q_{i+1}\right)\right)$$
	for $0 < i < k$, as well as the special cases
	$$\varphi^Q_{k} = \exists \vec A^k\left(\text{\normalfont Persistent}(k)\wedge \text{\normalfont Sat}_{\mathcal R}(k)\wedge \text{\normalfont Sat}_{\mathcal X}(k)\right)$$
	for leafs, and $$\varphi^Q = \varphi_0^Q = \exists \vec A^0\forall \vec X^0\left(\neg P^0\wedge \text{\normalfont Sat}_{\mathcal R}(0)\wedge(\text{\normalfont Valid}(0)\to \varphi_1^Q)\right)$$for the root.
	Example encodings of the above formulas are:
	\begin{align*}
		\text{\normalfont Valid}(n) = &\left(\bigvee\left\{X^n_\psi\:|\:\psi\in\mathcal X\right\}\right)\wedge\left(\bigwedge \left\{\neg(X^n_{\psi_1}\wedge X^n_{\psi_2})\:|\:\psi_1\neq\psi_2\in\mathcal X\right\}\right)\wedge\\&\left(\bigwedge\left\{\bigwedge\{X^i_\psi\to\neg X^n_\psi\:|\:i < n\}\:|\:\psi\in\mathcal X\right\}\right)\\
		\text{\normalfont Persistent}(n) = & \bigwedge\left\{A^{n-1}\to A^n\:|\: A \text{ prop. variable with }A^{n-1}\in\vec A^{n-1}, A^n \in\vec A^{n}\right\}\\
		\text{\normalfont Sat}_{\mathcal R}(n) = &\bigwedge\left\{\bigwedge A^n_i\to\bigvee B^n_j\:|\:\bigwedge A_i\to\bigvee B_j\mathcal R\right\}\\
		\text{\normalfont Sat}_{\mathcal X}(n) = & \bigwedge\left\{X^{n-1}_\psi\to (A^n\to B^n)\to C^{n-1}\:|\:\psi = (A\to B)\to C\in\mathcal X\right\}
	\end{align*}
\end{definition}

>>>>

A detailed account of the translation can be found in definition~\ref{def:fo-translation}. Finally, we benchmark our translation by translating all problems in the ILTP problem set~\cite{iltp} and applying the Vampire theorem prover~\cite{kovacs2013first} to the translated instances. While this method falls short of the absolute state of the art in general, under the right circumstances it is a reasonable approach. Since there is still a lot of room for improvement this is a hopeful first sign.

@@

\noindent All we have done is model Kripke semantics as a first-order theory, so we obtain:

@@

\begin{remark}
	It is well known that Kripke frames which are trees are complete with regard to intuitionistic logic. Our contribution is giving an explicit bound to height and degree that depends only on a simple property of the considered formula.
\end{remark}

@@

We can apply the previous results to give a direct embedding from intuitionistic into classical propositional logic. We know that $\varphi^{CH}$ is valid if and only if it is valid for structures that have as a domain the sequences without repetition over $\mathcal X$, ordered by the prefix-relation and $$s^I = \epsilon, f_\psi^I(x) = \begin{cases}
	x, &\text{ if $\psi$ occurs in $x$,}\\
	x\psi, &\text{ otherwise.}
\end{cases}$$
Now we can replace every $\forall$-quantifier in $\varphi^{CH}$ by enumerating all the ground terms, and then replace all distinct ground instances of relations by new propositional variables yielding a propositional formula. Note that the encoding of the partial order becomes redundant, so we drop it. We need to keep the persistency constraints. This gives us the following theorem:

\begin{theorem}\label{thm:prop-translation}
	Suppose we are given a formula as in Lemma~\ref{lemma:propositional-normal-form}, i.e. of the form
	$$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal S\right)\to P$$
	with flat $\mathcal R$ and implication $\mathcal S$. Let $\Lambda$ denote the set of sequences without repetition over $\mathcal X$. For every propositional variable $A$ and $\lambda\in\Lambda$ consider a new propositional Variable $A^\lambda$. Define $R^\#$ as the union over all sets
	$$
		\left\{\bigwedge_iA_i^\lambda\to\bigvee_jB_j^\lambda\:|\:\lambda\in\Lambda\right\}
	$$
	for $\bigwedge_iA_i\to\bigvee_jB_j\in\mathcal R$ and $\mathcal X^\#$ as the union over all sets
	$$
		\left\{\left(A^{\lambda\psi} \to A^{\lambda\psi}\right)\to C^\lambda\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	for $\psi = (A\to B)\to C\in\mathcal X$ as well as $P^\#$ as the union over all sets
	$$
		\left\{A^\lambda\to A^{\lambda\psi}\:|\:\lambda\in\Lambda, \psi\text{ does not occur in }\lambda\right\}
	$$
	 for each propositional variable $A$ occurring in $\varphi$, encoding persistency. Then
	$$
		\varphi^\# := \left(\bigwedge P^\#\wedge\bigwedge \mathcal R^\#\wedge\bigwedge\mathcal X^\#\right)\to P^\epsilon
	$$
	is classically valid if and only if $\varphi$ is intuitionistically valid.
\end{theorem}
%\begin{theorem}
%	$\bigwedge\mathcal S\to P$ is intuitionistically valid iff $\bigwedge\mathcal S^\#\to P^\emptyset$ is classically valid.
%\end{theorem}

@@

\begin{definition}
	For every propositional variable $A$ occurring in $\varphi$ and non-negative integer $n$ consider a new propositional variable $A^n$ and for every formula $\psi\in\mathcal X$ consider a new propositional variable $X_\psi^n$. Let $\vec A^n$ range over all propositional variables $A^n$ and $\vec X^n$ range over all $X_\psi^n$. Define
	\begin{itemize}
		\item {\normalfont Valid}$(n)$ which encodes that $\vec X^n_\psi$ represents a valid next element of a sequence without repetition, i.e. if we interpret the $\vec X^i$ as bit-vectors, then $\vec X^n$ has exactly one bit set to $1$, indicating the formula at the $n$-th position of the sequence, and it is at a position that is $0$ in $\sum\{\vec X^i\:|\:i < n\}$.
		\item {\normalfont Persistent}$(n)$ which encodes that the persistency condition holds.
		\item {\normalfont Sat}$_{\mathcal R}(n)$ encoding that the formulas in $\mathcal R$ hold at the $n$-th world of the path.
		\item {\normalfont Sat}$_{\mathcal X}(n)$ encoding that the formula in $\mathcal X$ that is represented by $\vec X^{n-1}$ holds.
	\end{itemize}
	Define$$\varphi^Q_{i} = \exists \vec A^i\forall \vec X^i\left(\text{\normalfont Persistent}(i)\wedge \text{\normalfont Sat}_{\mathcal R}(i)\wedge \text{\normalfont Sat}_{\mathcal X}(i)\wedge\left(\text{\normalfont Valid}(i)\to \varphi^Q_{i+1}\right)\right)$$
	for $0 < i < |\mathcal{X}|$, as well as the special cases
	$$\varphi^Q_{|\mathcal{X}|} = \exists \vec A^{|\mathcal{X}|}\left(\text{\normalfont Persistent}(|\mathcal{X}|)\wedge \text{\normalfont Sat}_{\mathcal R}(|\mathcal{X}|)\wedge \text{\normalfont Sat}_{\mathcal X}(|\mathcal{X}|)\right)$$
	for leafs, and $$\varphi^Q = \varphi_0^Q = \exists \vec A^0\forall \vec X^0\left(\neg P^0\wedge \text{\normalfont Sat}_{\mathcal R}(0)\wedge(\text{\normalfont Valid}(0)\to \varphi_1^Q)\right)$$for the root.
	Example encodings of the above formulas are:
	\begin{align*}
		\text{\normalfont Valid}(n) = &\left(\bigvee\left\{X^n_\psi\:|\:\psi\in\mathcal X\right\}\right)\wedge\left(\bigwedge \left\{\neg(X^n_{\psi_1}\wedge X^n_{\psi_2})\:|\:\psi_1\neq\psi_2\in\mathcal X\right\}\right)\wedge\\&\left(\bigwedge\left\{\bigwedge\{X^i_\psi\to\neg X^n_\psi\:|\:i < n\}\:|\:\psi\in\mathcal X\right\}\right)\\
		\text{\normalfont Persistent}(n) = & \bigwedge\left\{A^{n-1}\to A^n\:|\: A \text{ prop. variable with }A^{n-1}\in\vec A^{n-1}, A^n \in\vec A^{n}\right\}\\
		\text{\normalfont Sat}_{\mathcal R}(n) = &\bigwedge\left\{\bigwedge A^n_i\to\bigvee B^n_j\:|\:\bigwedge A_i\to\bigvee B_j\mathcal R\right\}\\
		\text{\normalfont Sat}_{\mathcal X}(n) = & \bigwedge\left\{X^{n-1}_\psi\to (A^n\to B^n)\to C^{n-1}\:|\:\psi = (A\to B)\to C\in\mathcal X\right\}
	\end{align*}
\end{definition}
